---
title: "HC Modeling"
author: "DataLAKE"
date: "2025-03-04"
output: html_document
---


# Preparation
## Housekeeping
Loading libraries and reading in data.
```{r}
library(dplyr)
library(data.table)   # for faster reading if desired
library(caret)        # for model training
library(nnet)         # neural network engine
library(RWeka)
```


```{r, read in data}

# Read Data (paths may differ)
application <- readRDS("application_train_cleaned.rds")
bureau <- fread("bureau.csv")
bureau_balance <- fread("bureau_balance.csv")

```

## Downsample
Creating a sample dataset to reduce training time. Adjusting for wildly imbalanced TARGET variable in the downsample.

```{r}

# Downsample with prejudice to fix target imbalance
# Split into majority/minority
minority <- application %>% filter(TARGET == 1)
majority <- application %>% filter(TARGET == 0)

set.seed(42)
# 70/30 split for downsample between majority/minority target classes
desired_ratio <- 70 / 30
desired_majority_size <- floor(desired_ratio * nrow(minority))
majority_downsampled <- majority %>%
  sample_n(size = desired_majority_size)

# Unsplit (making up words over here)
application_balanced <- bind_rows(minority, majority_downsampled)
table(application_balanced$TARGET)

```

## Aggregation
Aggregating auxillary tables so there is only one line per SK_ID number
```{r}
# Starting with just avg of months_balance
bureau_balance_agg <- bureau_balance %>%
  group_by(SK_ID_BUREAU) %>%
  summarise(
    MONTHS_BALANCE_MEAN = mean(MONTHS_BALANCE, na.rm = TRUE)
  )

```
Started with avg of months_balance, but might want to use min() so it's the most recent?

## Cleaning
Clean the auxillary tables before merging
```{r}
dim(bureau)
dim(bureau_balance)

str(bureau)
str(bureau_balance)

```

```{r}
bureau <- bureau %>%
  distinct()  # removes exact duplicates
bureau_balance <- bureau_balance %>%
  distinct()
# Function to get % of NA in each column
na_percentage <- function(df) {
  sapply(df, function(x) mean(is.na(x))) * 100
}

bureau_na_pct <- na_percentage(bureau)
bureau_balance_na_pct <- na_percentage(bureau_balance)

bureau_na_pct
bureau_balance_na_pct


```

```{r}
# (A) Convert to character
bureau$CREDIT_ACTIVE <- as.character(bureau$CREDIT_ACTIVE)
str(bureau$CREDIT_ACTIVE) 
# Check if it's "character" now

# (B) Replace NA with "Unknown"
bureau$CREDIT_ACTIVE[is.na(bureau$CREDIT_ACTIVE)] <- "Unknown"
table(bureau$CREDIT_ACTIVE, useNA = "ifany")

# (C) Re-factor
bureau$CREDIT_ACTIVE <- as.factor(bureau$CREDIT_ACTIVE)
str(bureau$CREDIT_ACTIVE)
table(bureau$CREDIT_ACTIVE, useNA = "ifany")


if ("CREDIT_ACTIVE" %in% names(bureau)) {
  bureau$CREDIT_ACTIVE <- as.character(bureau$CREDIT_ACTIVE)
  bureau$CREDIT_ACTIVE[is.na(bureau$CREDIT_ACTIVE)] <- "Unknown"
  bureau$CREDIT_ACTIVE <- as.factor(bureau$CREDIT_ACTIVE)
}

bureau$AMT_CREDIT_SUM[is.na(bureau$AMT_CREDIT_SUM)] <- median(bureau$AMT_CREDIT_SUM, na.rm = TRUE)
if("CREDIT_ACTIVE" %in% names(bureau)) {
  bureau$CREDIT_ACTIVE[is.na(bureau$CREDIT_ACTIVE)] <- "Unknown"
}
bureau <- na.omit(bureau)

```



```{r}
#write_csv(bureau, "bureau_cleaned.csv")
#write_csv(bureau_balance, "bureau_balance_cleaned.csv")

```




## Merging
Merging auxillary tables first.
```{r}
# Smish-smash let's join em up
bureau_joined <- bureau %>%
  left_join(bureau_balance_agg, by = "SK_ID_BUREAU")

bureau_summary <- bureau_joined %>%
  group_by(SK_ID_CURR) %>%
  summarise(
    count_records       = n(),
    avg_credit_sum      = mean(AMT_CREDIT_SUM, na.rm = TRUE),
    total_debt_sum      = sum(AMT_CREDIT_SUM_DEBT, na.rm = TRUE),
    max_overdue         = max(AMT_CREDIT_SUM_OVERDUE, na.rm = TRUE),
    avg_months_balance  = mean(MONTHS_BALANCE_MEAN, na.rm = TRUE)
  )
bureau_summary$avg_months_balance[is.na(bureau_summary$avg_months_balance)] <- 0

```

```{r, include = FALSE}
# Recalculate missing values after imputation
missing_after <- colSums(is.na(bureau_summary))
missing_after <- missing_after[missing_after > 0]  # Keep only columns still missing values (should be empty)

# Create a summary table for missing values after imputation
missing_after_table <- data.frame(
  Column = names(missing_after),
  Missing_Count = missing_after
)

# Print the table (should be empty if imputation worked)
print(missing_after_table)

# Double check for na values because I'm paranoid
sum(is.na(bureau_summary))  # Should return 0
```

Joining aux tables to main application table.
```{r}
# Smish-smash again, bureau tables to application table
merged_data <- application_balanced %>%
  left_join(bureau_summary, by = "SK_ID_CURR")

```

Because the join was a left to keep our sample large, there may be some IDs that not have missing data in columns. If intersect is much smaller, that means many ID numbers didn't match up.
```{r}
length(intersect(application_balanced$SK_ID_CURR, bureau_summary$SK_ID_CURR))
length(unique(application_balanced$SK_ID_CURR))
length(unique(bureau_summary$SK_ID_CURR))

```
Filling in missing data with 0
```{r}
merged_data <- merged_data %>%
  mutate(
    count_records      = if_else(is.na(count_records), 0, count_records),
    avg_credit_sum     = if_else(is.na(avg_credit_sum), 0, avg_credit_sum),
    total_debt_sum     = if_else(is.na(total_debt_sum), 0, total_debt_sum),
    max_overdue        = if_else(is.na(max_overdue), 0, max_overdue),
    avg_months_balance = if_else(is.na(avg_months_balance), 0, avg_months_balance)
  )

```



```{r, include = FALSE}
# Recalculate missing values after imputation
missing_after <- colSums(is.na(merged_data))
missing_after <- missing_after[missing_after > 0]  # Keep only columns still missing values (should be empty)

# Create a summary table for missing values after imputation
missing_after_table <- data.frame(
  Column = names(missing_after),
  Missing_Count = missing_after
)

# Print the table (should be empty if imputation worked)
print(missing_after_table)

# Double check for na values because I'm paranoid
sum(is.na(merged_data))  # Should return 0
```


## Data transformation
Removing irrelevant columns and transforming TARGET variable

```{r}
# Convert TARGET to factor for classification
merged_data$TARGET <- factor(merged_data$TARGET, levels = c(0,1))

# Remove IDs or any columns you don't want to feed to the model
merged_data <- merged_data %>%
  select(-SK_ID_CURR)

# Structure check
str(merged_data)

```


```{r}
sum(is.na(merged_data)) 
```


```{r}
saveRDS(merged_data, "merged_data.rds")
```

```{r}
impute_numeric_col <- function(df, col_name, strategy = "median") {
  # Check if the specified column exists
  if (!col_name %in% names(df)) {
    warning(paste("Column", col_name, "not found in df. Returning original df."))
    return(df)
  }
  # Check if it's numeric
  if (!is.numeric(df[[col_name]])) {
    warning(paste("Column", col_name, "is not numeric. Returning original df."))
    return(df)
  }
  
  # Determine the fill value based on the strategy
  fill_val <- NA
  if (strategy == "zero") {
    fill_val <- 0
  } else if (strategy == "min") {
    fill_val <- min(df[[col_name]], na.rm = TRUE)
  } else if (strategy == "max") {
    fill_val <- max(df[[col_name]], na.rm = TRUE)
  } else if (strategy == "median") {
    fill_val <- median(df[[col_name]], na.rm = TRUE)
  } else if (strategy == "mean") {
    fill_val <- mean(df[[col_name]], na.rm = TRUE)
  } else if (strategy == "sum") {
    fill_val <- sum(df[[col_name]], na.rm = TRUE)
  } else {
    stop("strategy must be one of 'zero', 'min', 'max', 'median', or 'mean'")
  }
  
  # Replace NA with the chosen fill value
  df[[col_name]][is.na(df[[col_name]])] <- fill_val
  
  return(df)
}

```
```{r}
#impute_numeric_col(merged_data, , 'zero')
```


# Modeling
## Partitioning
```{r}

# Partition the data for train/test 
set.seed(123)
trainIndex <- createDataPartition(merged_data$TARGET, p = 0.8, list = FALSE)
trainData  <- merged_data[ trainIndex, ]
testData   <- merged_data[-trainIndex, ]

# Suppose your target column is currently 0/1
trainData$TARGET <- factor(trainData$TARGET, 
                           levels = c(0, 1), 
                           labels = c("Class0", "Class1"))

# Double-check 
levels(trainData$TARGET)
# [1] "Class0" "Class1"

```


The neural network is really struggling with near-zero variance. Removing these columns
```{r}
library(caret)

# Identify columns with zero or near-zero variance
nzv_cols <- nearZeroVar(trainData, saveMetrics = TRUE)
nzv_cols
# This shows which columns are zeroVar or nearZeroVar

# If you want to remove only strictly zero-variance columns:
zero_var_cols <- rownames(nzv_cols[nzv_cols$zeroVar == TRUE, ])
trainData_clean <- trainData[, !(names(trainData) %in% zero_var_cols)]

# Or if you want to remove near-zero-variance too:
nzv_col_names <- rownames(nzv_cols[nzv_cols$nzv == TRUE, ])
trainData_clean <- trainData[, !(names(trainData) %in% nzv_col_names)]

```


## Model creation
Chose nnet as a simple, one layer model to evaluate efficacy of NN, will likely upgrade to other NN with multiple layers

### Build MLP models
Build an MLP model on MultilayerPerceptron()’s default setting on the training set. Evaluate the model performance on the training set and testing set.
```{r MLP model}
#Designate shortened name 'MLP' for the MultilayerPercentron ANN method in RWeka

MLP <- make_Weka_classifier("weka/classifiers/functions/MultilayerPerceptron")

#MLP model with default values

#MLP_default <- MLP(train_target ~ .,data = train_input)
#MLP_default
#summary(MLP_default)
```

```{r}
# Ensure we have an outcome formula
# We'll use all columns except "TARGET" as predictors
myFormula <- TARGET ~ .

# Training control: 5-fold CV, measure ROC
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

# Train a neural network
set.seed(123)
MLP_model <- caret::train(
  myFormula,
  data = trainData,
  method = "MLP",           # uses the nnet package under the hood
  trControl = ctrl,
  metric = "ROC",            # optimize area under ROC
  preProcess = c("center","scale", "nzv"),  # standardize numeric predictors, remove near zero variance
  tuneLength = 5,            # try 5 combos of hyperparameters
  trace = FALSE              # suppress training logs
)

# Inspect the model
MLP_model

```

## Evaluation
```{r}
pred_probs <- predict(nn_model, newdata = testData, type = "prob")  # predicted probabilities
pred_class <- predict(nn_model, newdata = testData, type = "raw")   # predicted classes

# 1. Make sure your reference (testData$TARGET) is a factor with levels "0","1"
testData$TARGET <- factor(testData$TARGET, levels = c("0","1"))

# 2. Make sure pred_class is also a factor with the same levels
pred_class <- factor(pred_class, levels = c("0","1"))

# 3. Then confusionMatrix should work
confMat <- confusionMatrix(pred_class, testData$TARGET, positive = "1")
confMat


# ROC curve, AUC, etc.
library(pROC)
roc_obj <- roc(response = testData$TARGET, predictor = pred_probs[,"Class1"])
auc(roc_obj)

```


```{r}

```













